---
title: 'AML vs ALL'
author: "Jenny Smith"
date: "November 13, 2018"
output: html_document
---



```{r setup}
library(knitr)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=50),tidy=TRUE, fig.align='center', fig.height=5, fig.width=8)
knitr::opts_knit$set(root.dir = '/fh/fast/meshinchi_s/workingDir/TARGET/AML_TARGET/RNA/mRNAseq/analysis/2018.11.13_AML_vs_ALL/')
options(stringsAsFactors = FALSE)
```


```{r message = FALSE, warning=FALSE}
library(stringr)
library(magrittr)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
getwd()
```


#Define Functions


```{r}
matchCompartment <- function(gene, df){
  #keywords: extracellular, plasma membrane, transmembrane
  library(dplyr)
  
  
  regex <- paste0("\\b", gene,"\\b")
    
  comp <- df %>%
      filter(grepl(regex, V2, ignore.case = TRUE)) %>%
      filter(grepl("extracellular|plasma membrane|transmembrane|Cell periphery", V4, ignore.case = TRUE)) %>%
      dplyr::select(V4)
    
  res <- paste(unlist(unique(comp)),collapse="; ")
  
  
  return(res)
}

```




#Read in the Clinical Data


```{r}
CDE.0531 <- read.csv("~/reference_mapping-files/TARGET_AML_current_asof_june30_2016_UPDATED_CLEAN_4.19.18.csv")

rownames(CDE.0531) <- CDE.0531$TARGET.USI.1
head(CDE.0531[,1:5])
dim(CDE.0531)
```

```{r}
pheno.data <- read.delim("/fh/fast/meshinchi_s/workingDir/TARGET/NormalTissue_GTEX/Clinical/TcgaTargetGTEX_phenotype.txt")

head(pheno.data)
dim(pheno.data)
```



#Read in the Expression Data

NOTE: this dataset was already converted into linear scale. NOT log2(TPMs) anymore. 
```{r}
TPMs <- get(load("/fh/fast/meshinchi_s/workingDir/TARGET/NormalTissue_GTEX/RNA/mRNAseq/level3/gene/2016Sept_UCSC_Illumina_data/Rdata/TcgaTargetGtex_rsem_dupGenesRemoved_tpm.RData"))

head(TPMs[,1:5])
dim(TPMs) #58531 19261
```


#Read in the Gene Annotations

```{r}
compartments <- read.delim("~/RNA_seq_Analysis/0000.00.02_Reference_GeneInfo/human_compartment_knowledge_full.tsv", sep = "\t",as.is = TRUE, stringsAsFactors = FALSE, header = FALSE)
  
dim(compartments)
```



#Select Samples of Interest 

```{r}
data.types <- pheno.data %>%
  group_by(detailed_category) %>%
  summarize(Number=n()) 

```


```{r}
pheno.AML_ALL <- pheno.data %>%
  filter(X_study != "TCGA") %>%
  filter(grepl("Acute Lymphoblastic Leukemia|Acute Myeloid Leukemia|Whole Blood", detailed_category)) %>%
  mutate(sample.colname=gsub("-", "\\.", sample)) %>%
  mutate(Group=case_when(
    grepl("Whole Blood", detailed_category) ~ "NormalBlood",
    grepl("Lympho", detailed_category) & grepl("Recurrent", X_sample_type) ~ "ALL_Relapse",
    grepl("Lympho", detailed_category) & ! grepl("Recurrent", X_sample_type) ~ "ALL_Diagnostic",

    grepl("Myeloid", detailed_category) & grepl("Recurrent", X_sample_type) ~ "AML_Relapse",
    grepl("Myeloid", detailed_category) & ! grepl("Recurrent", X_sample_type) ~ "AML_Diagnostic"))
  

head(pheno.AML_ALL)
dim(pheno.AML_ALL) #759   7
```

```{r}
table(pheno.AML_ALL$detailed_category, pheno.AML_ALL$X_study)
table(pheno.AML_ALL$Group)

write.csv(table(pheno.AML_ALL$Group),"GroupSizes.csv", row.names=FALSE)
```


#Compare the Expression of Genes Across AML and ALL

```{r}
TPMs.m <- TPMs %>%
  rownames_to_column("Gene") %>%
  select(Gene,pheno.AML_ALL$sample.colname) %>%
  filter(rowSums(select_if(., is.numeric)) > 0) %>% #remove non-expressed genes only 
  
  gather(Sample, TPM, -Gene) %>%
  left_join(., pheno.AML_ALL, by=c("Sample"="sample.colname")) %>%
  
  mutate(Group=case_when(
    grepl("Whole Blood", detailed_category) ~ "NormalBlood",
    grepl("Lympho", detailed_category) & grepl("Recurrent", X_sample_type) ~ "ALL_Relapse",
    grepl("Lympho", detailed_category) & ! grepl("Recurrent", X_sample_type) ~ "ALL_Diagnostic",

    grepl("Myeloid", detailed_category) & grepl("Recurrent", X_sample_type) ~ "AML_Relapse",
    grepl("Myeloid", detailed_category) & ! grepl("Recurrent", X_sample_type) ~ "AML_Diagnostic"))

head(TPMs.m[,])
dim(TPMs.m) #50918 by  760 (about 8,000 genes not expressed at all)
```

There appears to be an oddly great amount of normals with low expression across tons of genes. 
I am starting to wonder if this dataset is really any good at all.... 

```{r}
normals <- TPMs.m %>%
  filter(Group=="NormalBlood") %>%
  
  #for each gene, finds the ones with 0.5 TPMs in all normals
  #and the mean for normals must be less than 0.5 TPM across  all samples
  group_by(Gene) %>%
  mutate(Low.Expn=sum(TPM <= 0.5) >= n()*0.8) %>%
  mutate(Mean.Expn=mean(TPM)) %>%
  mutate(Low.Mean=Mean.Expn <= 0.5) %>%
  ungroup() %>%
  
  # #filter only genes that satisfy the low expression criteria
  filter(Low.Expn==TRUE, Low.Mean==TRUE) %>%
  arrange(desc(Mean.Expn))

head(normals)

#35,987 out of 50,000 genes have VERY low expression... Would really expect this is to be fewer genes. like 9-10K at most...
length(table(normals$Gene)) 
```

```{r}
TPMs.comp <- TPMs.m %>%
  
  #start with looking for genes lowly expressed in normals
  filter(Gene %in% normals$Gene) %>%
  
  #Identify genes with at least 25% of patients are high expressors, and mean across all samples is high. 
  group_by(Group,Gene) %>%
  mutate(High.Expn=sum(TPM >= 10) >= n()*.25) %>%
  ungroup() 
  

options(scipen=999)
head(TPMs.comp)
dim(TPMs.comp)
length(unique(TPMs.comp$Gene))
```


```{r}
GOI <- TPMs.comp %>%
  #select genes that have high expression in ALL and AML diagnostic samples 
  filter(grepl("Diagnostic", Group)) %>%
  select(Gene,Group,High.Expn) %>%
  arrange(Gene) %>%
  
  #de-duplicate the dataframe since I want only 2 entries per gene, 1 for AML as T/F and 1 for ALL as T/F
  group_by(Gene) %>%
  filter(!duplicated(Group)) %>%
  ungroup() %>%
  
  #Select genes with hgih expression in BOTH leukemias
  spread(Group,High.Expn)  %>%
  filter(AML_Diagnostic==TRUE, ALL_Diagnostic==TRUE)

  

head(GOI)
dim(GOI) #only 28 genes... 
```

  # select(-primary.disease.or.tissue, 
  #        -X_primary_site, -X_study, -primary.disease.or.tissue, -High.Expn, -X_gender) %>%
vars(AC002454.1:ZBTB10)

Percent.Expressors=round((sum(. >= 10, na.rm = T)/n())*100, digits = 2))
Number.Samples=n()

  
```{r}
TPMs.GOI.mat <- TPMs.comp %>%
  filter(Gene %in% GOI$Gene) %>%
  select(Sample,Gene,Group,TPM) %>%
  spread(Gene,TPM) 


head(TPMs.GOI.mat)
dim(TPMs.GOI.mat)

# write.csv(TPMs.GOI.mat, "Genes_High_in_AML_ALL_TOIL_Reprocessed_TPMs.csv", row.names = FALSE)
```

```{r}
TPMs.GOI <- TPMs.comp %>%
  filter(Gene %in% GOI$Gene) %>%
  spread(Gene,TPM) %>%
  
  group_by(Group) %>%
  summarize_at(vars(AC002454.1:ZBTB10), funs(
                               Max.TPM=max(., na.rm = T), 
                               Median.TPM=median(., na.rm = T), 
                               Mean.TPM=mean(., na.rm = T), 
                               
                               Number.Expressors=sum(. >= 10, na.rm = T), 
                               Number.In.Group=n(), 
                               Percent.Expressors=round((sum(. >= 10, na.rm = T)/n())*100, digits = 2))) %>%
  ungroup() %>%
  
  gather(Stat,Results,matches("Max|Mean|Median|Number|Percent")) %>%
  mutate(Gene=gsub("^(.+)_[MNP].+", "\\1", Stat),
         Stat=gsub("^.+_([MNP])", "\\1", Stat)) %>%
  spread(Stat,Results) %>%

  select(Group,Gene, matches("Number"), matches("Percent") ,matches("max|mean|median")) %>%
  arrange(Gene) %>%
  
  rowwise() %>%
  mutate(Cellular.Compartment=matchCompartment(gene=Gene,df=compartments)) %>%
  ungroup()



dim(TPMs.GOI)
head(TPMs.GOI)
# write.csv(TPMs.GOI,"Genes_High_in_AML_ALL_TOIL_Reprocessed_SummaryStats.csv", row.names = FALSE)
```



#Session Information 

```{r}
sessionInfo()
```

